\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage[numbers]{natbib}

\title{When Does Deliberation Improve an LLM Judge?}
\author{Working research draft}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
Reasoning models spend substantially more inference-time computation before
answering difficult questions, but an evaluator does not always have a
verifiable objective that tells it when its reasoning is complete. We study
how native inference-time deliberation changes pairwise decisions as the
decision criterion moves from exact and factual verification to consensus
judgments, multi-attribute preferences, and open-ended human-versus-LLM
evaluation. Our initial Qwen3-8B results suggest a domain-dependent tradeoff:
thinking corrects some factual mistakes, while evaluative decisions consume
more tokens, fail to terminate more often, and do not reliably become more
branch-consistent. The planned experiments cross token budgets with
within-pair criterion interventions, then trace decision evidence with fixed
held-out activation decoders. The intended contribution is a characterization
of when additional inference-time computation functions as useful
verification and when it becomes prolonged search without a determinate
stopping rule.
\end{abstract}

\section{The Paper Story}

Chain-of-thought is not uniformly useful. Broad evaluations find its clearest
benefits on mathematical and symbolic tasks \citep{sprague2024tocot}, while
other work identifies tasks on which verbal deliberation can reduce
performance \citep{liu2024mindstep}. Native reasoning models make this question
more consequential because they allocate hundreds or thousands of tokens
without an external verifier deciding when enough computation has occurred.
Qwen3 is a useful substrate because the same post-trained checkpoint exposes
native thinking and non-thinking modes \citep{qwen2025qwen3}.

LLM judging is a particularly sharp test case. Some judge decisions have
checkable criteria, such as deciding which proposed answer is factually
correct. Others ask the model to aggregate helpfulness, correctness, clarity,
style, and values without specifying a total ordering. Existing judge research
shows that such decisions are sensitive to response order, prompt framing,
verbosity, and model identity \citep{zheng2023judging,wang2023position,
panickssery2024selfrecognition}. Additional deliberation could improve these
decisions by integrating more evidence. It could instead generate a longer
rationale around a choice for which no unique verification signal exists.

Our central hypothesis is therefore not that reasoning is generally harmful.
It is that the return to inference-time computation depends on the
determinacy of the decision criterion. Exact tasks provide a recognizable
success condition. Plural judgments may not.

\section{Initial Evidence}

In the first Qwen3-8B scout, native thinking improved majority decisions on ARC
Challenge and TruthfulQA, including wrong-to-correct transitions without
corresponding correct-to-wrong majority transitions. The same condition
generated much longer traces on HelpSteer2 tradeoffs, D4 human-versus-LLM
pairs, and ETHICS Commonsense. It also produced substantially more
budget-saturated invalid outputs. Branch agreement did not improve reliably.

This is suggestive but insufficient. The current result confounds criterion
determinacy, ordinary difficulty, and a finite output budget. It also measures
natural termination rather than the decision the model would make if stopped
at a controlled prefix. The next experiment addresses both limitations.

\section{Experiment 1: Budgeted Deliberation}

We construct a progression from mathematical verification and logical
deduction through factual and consensus decisions to dominated preferences,
attribute-level tradeoffs, overall-quality tradeoffs, and ecological
human-versus-LLM judgments. For every sampled reasoning branch, we generate one
maximal native thinking trace. We then truncate that same trace at fixed token
budgets, close the reasoning block, and score an A/B verdict. This design
separates the information gained from additional reasoning from independent
sampling noise across budget conditions.

The primary result is a budget-by-task interaction. For exact and factual
tasks, useful deliberation should increase target accuracy and then plateau.
For underspecified judgments, the criterion-uncertainty hypothesis predicts
longer natural traces and lower termination without comparable improvements in
order consistency, branch agreement, or defensible target performance.

\section{Experiment 2: Criterion Rescue}

Task categories alone cannot show that the criterion causes the difference.
We therefore hold HelpSteer2 and D4 response pairs fixed while changing the
decision rule. The model judges the same pair by overall quality, correctness,
helpfulness, coherence or clarity, and an explicit weighted rubric.
HelpSteer2's attribute annotations provide criterion-specific proxy targets.
D4 remains unlabeled but supports termination, consistency, and authorship
analyses.

A criterion-rescue result would show that explicit decision rules shorten
reasoning, reduce budget saturation, or improve stability relative to an
overall-quality judgment on the same texts. This would distinguish criterion
uncertainty from subject matter and text difficulty.

Our first evidence collector makes the HelpSteer2 comparison fully matched.
It samples Pareto-dominant, single-attribute, tradeoff, and annotation-near-tie
pairs, then presents every pair under every criterion and in both response
orders. The judge may answer A, B, or C, where C is explicitly defined as tied
or underdetermined under the stated rule. Correctness in this experiment means
compliance with that rule and the corresponding HelpSteer2 attribute proxy. It
does not mean that HelpSteer2 defines a universal ranking of response quality.
The strongest outcome is therefore criterion-contingent switching with order
invariance: the verdict changes when the annotated target changes, remains
stable when it should not change, and uses C selectively rather than as an
escape response.

\section{Experiment 3: Decision Formation}

Visible reasoning is not assumed to faithfully report the computation that
caused the answer \citep{lanham2023faithfulness}. The mechanistic analysis asks
when final-choice and criterion information become linearly recoverable from
residual states. Unlike point-specific probes, the paper-facing analysis trains
a decoder only on endpoint states, selects it on separate pairs, freezes it,
and applies it across earlier states on held-out pairs. This creates a
comparable evidence scale across generated-token positions.

The descriptive outcomes are evidence emergence, persistent commitment,
revisions, confidence velocity, path efficiency, and branch convergence. A
particularly relevant pattern would be stable choice evidence before stable
criterion evidence on underspecified judgments. That pattern remains
associational until replay interventions with sign-reversed and random
orthogonal controls change final verdicts.

\section{Experiment 3a: Criterion Representation and Use}

The matched HelpSteer2 scout suggests a more specific failure than generic
evaluative difficulty. Qwen3-8B usually preserves a semantic preference when
the requested criterion changes, even while extended reasoning raises
confidence and sometimes increases order stability. We therefore distinguish
criterion representation, criterion-conditioned target construction, and
choice control.

On a held-out HelpSteer2 sample, each response pair supplies two criteria whose
targets either oppose one another, differ by tie versus choice, or agree. A
staged protocol elicits 128 tokens of analysis, then either preserves, reminds,
switches, withholds, or neutrally restates the decision rule before a further
384 tokens. Hidden A/B/C readouts measure whether a criterion update changes
the model's current verdict before and after additional reasoning.

The activation artifact replays these exact sampled prefixes. Point-level
labels identify the active criterion, its implied semantic target, and the
eventual semantic choice. Pair-held-out endpoint decoders are frozen and
applied across time. This yields three separable outcomes: the criterion fails
to update, the criterion updates without a target update, or criterion and
target update while choice remains anchored. Only the third pattern directly
supports a rationalization-like dissociation.

The confirmatory readout is taken at the exact forced-verdict boundary after
closing the reasoning block and appending \texttt{FINAL:}. We distinguish this
state from the last free-generation token. Decoder hyperparameters are selected
on frozen fit and selection pairs, while temporal estimates are cross-fitted
and bootstrapped by source pair so that response-order variants, conditions,
and sampled branches are not treated as independent observations.

We first calibrate the same readout on completed factual and verification
traces from ARC Challenge, BBH logical deduction, GSM8K, MATH-500, and
TruthfulQA at matched token budgets. This comparison does not turn probe
accuracy into a universal quality score. It establishes the within-model
recoverability and temporal-transfer pattern obtained when the decision
criterion has an externally checkable target.

\section{Experiment 3b: Criterion-State Patching}

Descriptive decoding does not establish control. For the same pair, order, and
sampled branch, we subtract the reminder-condition residual state from the
switch-condition state. Adding that within-pair criterion difference to a
reminder replay tests whether criterion-induced state changes are sufficient
to redirect the verdict. Negative, same-target placebo, shuffled-pair, and
presentation-order controls bound nonspecific perturbation effects.

The causal claim remains local to the tested layer, point, model, and replay
protocol. A successful patch shows criterion-state influence over the verdict;
an accurate decoder without a successful patch shows representation without
demonstrated causal use.

\section{Experiment 3c: Locked Operationalization Confirmation}

The initial 72 criterion-switch pairs are treated as development data. We
freeze the prompt construction and readout protocol before selecting 24 new
HelpSteer2 pairs: nine choice-to-choice conflicts, nine transitions from an
initial criterion tie to a criterion-resolved choice, and six same-target
controls. A blinded human audit checks both active criteria in both response
orders, for 96 criterion-specific judgments.

The confirmatory manipulation crosses whether the updated criterion is
available before or after a 128-token initial analysis with whether the model
also receives the two criterion-specific HelpSteer scores. The scores make the
criterion comparison directly computable but do not state the A/B/C verdict.
A separate late explicit-target condition supplies a downstream ceiling on 12
conflict pairs.

This factorial distinguishes three explanations. A general evidence benefit
indicates that the criterion-only prompt did not make the target sufficiently
computable. A selective loss of that benefit after prior commitment indicates
that answer formation gates later criterion use. Failure even when the
criterion target is supplied directly places the bottleneck downstream of
target construction. We reserve ``rationalization-like'' for the
commitment-dependent pattern and require causal replay before claiming that
the generated explanation preserves rather than produces the verdict.

\paragraph{Current confirmatory pattern.}
The locked behavioral sample does not show a meaningful early-versus-late
commitment penalty. Criterion-only forced target adoption is approximately
$0.59$ when the updated criterion is early and $0.53$ when it is late.
Supplying criterion-specific score evidence raises both conditions to
approximately $0.99$, and a late explicit target reaches $1.00$. The strongest
current interpretation is therefore that the model can respond to an
operationalized criterion target but does not reliably construct that target
from criterion language alone. This supports awareness or representation
without reliable operationalization; it does not yet support
commitment-driven rationalization.

\paragraph{Mechanistic adjudication.}
We freeze the development-selected readout layers before opening the
confirmatory activations. Pair-held-out probes measure the active criterion,
criterion target, current choice, and final choice at the exact
\texttt{FINAL:} boundary. The causal test patches three matched residual
differences: early criterion minus late criterion after the phase-one analysis,
late score evidence minus late criterion immediately after the phase-two
update, and late explicit target minus late criterion at the same point.
Positive patches are compared with sign-reversed, norm-matched random
orthogonal, shuffled same-target, shuffled opposite-target, and
same-target-transition controls. A selective increase in target probability
or target logit margin would show that the missing operationalized state is
sufficient to influence the verdict. Strong decoding without a selective
patch effect would instead localize the failure to downstream use.

\section{Experiment 3d: Does Structured Reasoning Compile the Rubric?}

The operationalization gap may arise because reasoning post-training teaches
models to continue familiar computational trajectories, while a natural
language criterion alone does not specify a sufficiently familiar procedure.
We test this account without changing the responses, criteria, targets, or
human audit. On the same locked 24 pairs, free CoT is compared with two
four-step scaffolds. The criterion-operational scaffold translates the active
rule into observable tests, applies the same tests to each option, and then
compares the criterion-specific assessments. The generic scaffold matches the
step count and approximate instruction length but instead requests balanced
summaries and surface comparison without compiling the criterion.

Score evidence remains an operationalization ceiling, and the existing
explicit-target condition remains a downstream ceiling. A deduplicated
direct-answer condition measures whether generated reasoning helps beyond the
model's immediate criterion-conditioned readout. The primary outcome is
order-consistent target adoption across both response orders, followed by
forced target adoption, natural validity, confidence, and budget saturation.
Checkpoint readouts determine whether any scaffold advantage emerges during
the initial 128-token analysis or only during the subsequent 384-token
continuation. A branch-zero activation replay applies the previously frozen
criterion, target, current-choice, and final-choice decoders to all four main
conditions. This tests whether behavioral rescue is accompanied by a stronger
criterion-conditioned target representation rather than only a different
surface rationale.

A criterion-scaffold advantage over both free CoT and generic structure would
support the hypothesis that an imposed computation sequence helps compile
rubric language into a usable target. Improvement over free CoT but not generic
structure would instead implicate generic attention or decomposition. If
scores remain effective while both scaffolds fail, the stronger conclusion is
that criterion-specific numerical evidence supplies a target-like state that
natural language reasoning procedures do not reliably construct.

We separate failure of the intervention from failure to follow its
instructions. Phase-one rationales are scored for explicit step structure and
for substantive criterion operationalization: criterion reference,
operational tests, symmetric application and textual grounding for both
options, criterion-specific comparison, and withholding of a premature final
answer. Pair-clustered within-pair analyses test whether stronger adherence is
associated with forced and order-consistent target adoption. Because
adherence is a post-treatment trace property, this analysis is a manipulation
check and descriptive association rather than a causal mediation result. A
manually reviewed stratified sample is required before treating automatic
adherence labels as a paper result.

\section{Experiment 3e: Computationally Enforced Deliberation}

The prompted scaffold increased criterion-oriented rationale content, but the
initial 128-token analysis saturated in every condition and did not complete
the requested four-step format. We therefore separate instruction following
from computational sequencing. On the same locked pairs, a criterion pipeline
first generates tests without seeing either option, freezes those tests,
assesses each option in a separate call, compares the isolated assessments,
and only then forms a verdict. A generic pipeline matches the number of calls
and token allocation using neutral planning, isolated summaries, and generic
comparison. Long free CoT and long single-prompt criterion scaffolding control
for the increased 1536-token analysis budget.

Freezing tests before option exposure directly limits post-hoc
criterion construction. The principal contrast is enforced criterion
reasoning against long free CoT, long prompted structure, and staged generic
decomposition on order-consistent target adoption. A selective rescue would
show that the missing operation can be supplied through an external
computation graph. A null result with low per-stage saturation would instead
show that neither additional tokens nor mandatory decomposition is sufficient
to turn criterion awareness into reliable criterion use.

\section{Replication and Scope}

The complete design is discovered and frozen on Qwen3-8B. Qwen3-30B-A3B is the
preferred scale replication because it supports the same native
thinking/non-thinking interface while activating relatively few parameters.
Qwen3-32B supplies a dense same-family check. Phi-4-reasoning-plus can provide
a cross-family replication after the behavioral estimands are fixed.
Reasoning-specialized checkpoints are post-trained systems, so the main
comparison is inference mode and criterion structure rather than a claim that
pretrained base models expose the same reasoning policy.

\section{Claim Ledger}

\begin{description}[leftmargin=3.2cm,style=nextline]
\item[Motivated]
Thinking benefits some factual controls but creates more nontermination on
evaluative tasks in Qwen3-8B.
\item[Not established]
The effect is caused by criterion uncertainty rather than finite budget,
difficulty, or prompt format.
\item[Required]
Budget curves, criterion rescue, one approximately 30B replication, and fixed
held-out trajectory decoders.
\item[Criterion-use gate]
The active criterion and its implied target must be decoded on held-out pairs.
A rationalization-like claim additionally requires persistent choice evidence
despite criterion or target updating.
\item[Causal gate]
Only controlled activation replay can support a claim that an early decoded
direction influences the final verdict.
\end{description}

\bibliographystyle{plainnat}
\bibliography{references}
\end{document}
